---
title: "Bayesian Urban Air Quality Forecasting & Health Risk Analysis"
subtitle: "MSc Statistics with Data Science — Portfolio Project"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  warning     = FALSE,
  message     = FALSE,
  fig.align   = "center",
  cache       = TRUE,
  cache.path  = "reports/cache/"
)

library(tidyverse)
library(lubridate)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(scales)
```

---

# Executive Summary

This report presents a complete statistical analysis of urban air quality in London, 
examining the relationship between PM2.5 particulate pollution and respiratory 
hospital admissions. The analysis spans six statistical and machine learning 
methodologies, directly corresponding to core modules in the MSc Statistics with 
Data Science programme at the University of Edinburgh.

**Key Findings:**

- PM2.5 concentrations show strong seasonality (winter peaks) and a long-term 
  downward trend across all monitoring sites
- Each 10 µg/m³ increase in PM2.5 is associated with approximately a 1.5% increase 
  in same-day respiratory admissions (Negative Binomial GLM)
- Bayesian hierarchical modelling confirms this effect with high posterior probability, 
  and provides richer uncertainty quantification than classical methods
- The October 2021 ULEZ expansion caused an estimated immediate reduction of 
  2–4 µg/m³ in inner London PM2.5 (Interrupted Time Series analysis)
- ARIMA time series models provide good short-term forecasting performance 
  (RMSE < 3 µg/m³ on 90-day test set)
- XGBoost marginally outperforms ARIMA on prediction, but at the cost of 
  interpretability

---

# Introduction

Air pollution is the largest environmental health risk in the UK, responsible for 
approximately 40,000 premature deaths annually (Public Health England, 2023). 
Fine particulate matter (PM2.5 — particles smaller than 2.5 micrometres) is 
particularly harmful, penetrating deep into the lungs and bloodstream.

The Ultra Low Emission Zone (ULEZ) in London represents one of the most significant 
air quality policy interventions in recent UK history. Understanding its causal 
impact requires rigorous statistical methodology — exactly the kind of real-world 
problem this analysis addresses.

## Objectives

1. Characterise the spatial and temporal patterns of London air quality (2018–2023)
2. Quantify the association between PM2.5 and respiratory hospital admissions
3. Forecast future pollution levels using time series methods
4. Estimate the causal effect of the ULEZ expansion using interrupted time series
5. Compare statistical and machine learning approaches to prediction

## Data Sources

```{r data-sources-table, echo=FALSE}
data.frame(
  Dataset = c("Air Quality", "Health Outcomes", "Intervention"),
  Source  = c("UK DEFRA AURN via openair R package",
               "Synthetic (generated with published dose-response parameters)",
               "London ULEZ expansion, 25 October 2021"),
  Coverage = c("4 London sites, 2018–2023",
                "Daily respiratory admissions, 2018–2023",
                "Inner vs outer London comparison"),
  Variables = c("PM2.5, NO2, O3, temperature, wind",
                 "Daily count, city-wide PM2.5 exposure",
                 "Pre/post binary + site treatment status")
) %>%
  kable(caption = "Data Sources Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

# Exploratory Data Analysis

```{r load-data}
# Load all processed data
pollution     <- readRDS("../data/pollution_clean.rds")
health        <- readRDS("../data/health_data.rds")
eda_summary   <- readRDS("../data/eda_summary.rds")

cat("Pollution observations:", nrow(pollution))
cat("\nDate range:", format(min(pollution$date)), "to", format(max(pollution$date)))
cat("\nSites:", paste(unique(pollution$site_code), collapse = ", "))
```

## PM2.5 Time Series

```{r pm25-timeseries, fig.cap="Daily PM2.5 concentrations at four London sites with LOESS smoothing."}
pollution %>%
  filter(!is.na(pm2.5)) %>%
  ggplot(aes(x = date, y = pm2.5, colour = site_code)) +
  geom_line(alpha = 0.3, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.15, se = FALSE, linewidth = 0.9) +
  geom_vline(xintercept = as.Date("2021-10-25"), 
             linetype = "dashed", colour = "red", linewidth = 0.8) +
  annotate("text", x = as.Date("2021-11-15"), y = 45,
           label = "ULEZ (Oct 2021)", colour = "red", size = 3.5, hjust = 0) +
  facet_wrap(~site_code, ncol = 2) +
  scale_colour_viridis_d() +
  labs(title = "Daily PM2.5: Four London Sites (2018–2023)",
       x = NULL, y = "PM2.5 (µg/m³)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

## Missing Data Analysis

```{r missing-data}
missing_pct <- pollution %>%
  group_by(site_code) %>%
  summarise(
    n_total   = n(),
    n_missing = sum(is.na(pm2.5)),
    pct_miss  = round(n_missing / n_total * 100, 2)
  )

missing_pct %>%
  kable(caption = "Missing PM2.5 data by site",
        col.names = c("Site", "Total days", "Missing days", "% Missing")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Missing data in air quality monitoring arises from sensor maintenance, 
power outages, and calibration periods. The pattern is **Missing Not At Random (MNAR)** 
— outages cluster during maintenance windows. We address this using **Multiple 
Imputation by Chained Equations (MICE)** with predictive mean matching, preserving 
the empirical distribution of observed values.

## Seasonal Patterns

```{r seasonal-patterns, fig.cap="PM2.5 by season across monitoring sites."}
pollution %>%
  filter(!is.na(pm2.5)) %>%
  mutate(season = factor(season, levels = c("Winter","Spring","Summer","Autumn"))) %>%
  ggplot(aes(x = season, y = pm2.5, fill = season)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
  facet_wrap(~site_code) +
  scale_fill_manual(values = c("Winter"="#5B8DBE","Spring"="#4CAF50",
                                "Summer"="#FFC107","Autumn"="#FF7043")) +
  labs(title = "PM2.5 by Season", x = NULL, y = "PM2.5 (µg/m³)") +
  theme_minimal(base_size = 11) + theme(legend.position = "none")
```

Winter PM2.5 is consistently higher due to:

1. **Residential heating** — wood and solid fuel burning
2. **Atmospheric inversions** — cold air traps pollution near the surface
3. **Reduced photochemical dispersion** — less UV light in winter

## WHO Guideline Exceedances

```{r who-exceedances}
who_exceedances <- pollution %>%
  filter(!is.na(pm2.5)) %>%
  group_by(site_code, year) %>%
  summarise(
    annual_mean = round(mean(pm2.5), 1),
    days_over   = sum(pm2.5 > 15),
    pct_over    = round(mean(pm2.5 > 15) * 100, 1),
    .groups = "drop"
  )

who_exceedances %>%
  pivot_wider(names_from = site_code, values_from = c(annual_mean, pct_over)) %>%
  kable(caption = "Annual mean PM2.5 and % days exceeding WHO guideline (15 µg/m³)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Generalised Regression Models

```{r load-glm}
glm_results <- readRDS("../data/glm_results.rds")
model_data  <- glm_results$model_data
```

## Model Specification

We model daily respiratory admissions $Y_t$ as a function of pollution exposure:

$$Y_t \sim \text{NegBin}(\mu_t, \theta)$$

$$\log(\mu_t) = \beta_0 + \beta_1 \cdot \text{PM2.5}_{t} + \beta_2 \cdot \text{PM2.5}_{t-1} + \beta_3 \cdot \text{PM2.5}_{t-2} + \beta_4 \cdot \text{PM2.5}_{t-3} + \gamma \cdot \text{Season}_t + \delta \cdot \text{Weekday}_t$$

The **Negative Binomial** family is chosen over Poisson because count health data 
is typically **overdispersed** (variance > mean). We test this formally:

```{r overdispersion-test}
m1 <- glm_results$m1_poisson
dispersion_ratio <- m1$deviance / m1$df.residual
cat("Dispersion ratio (Residual deviance / df):", round(dispersion_ratio, 3))
cat("\nRatio > 1.5 indicates overdispersion requiring Negative Binomial")
```

## Model Comparison

```{r model-comparison-table}
glm_results$model_comparison %>%
  kable(caption = "GLM Model Comparison — lower AIC = better fit",
        digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1, bold = TRUE, background = "#e8f4f8")  # highlight best model
```

## Rate Ratios

```{r rr-plot, fig.cap="Forest plot of Rate Ratios for pollution effects on respiratory admissions."}
glm_results$coefs %>%
  ggplot(aes(x = rr, y = reorder(term_label, rr))) +
  geom_vline(xintercept = 1, colour = "gray50", linetype = "dashed") +
  geom_errorbarh(aes(xmin = rr_low, xmax = rr_high), height = 0.25,
                 colour = "#5B8DBE", linewidth = 1.2) +
  geom_point(size = 3.5, colour = "#FF7043") +
  labs(title = "Rate Ratios: Effect of Pollution on Respiratory Admissions (Negative Binomial GLM)",
       subtitle = "Rate Ratios > 1 = increased risk. Error bars = 95% CI.",
       x = "Rate Ratio (RR)", y = NULL) +
  theme_minimal(base_size = 12)
```

---

# Time Series Analysis

```{r load-ts}
ts_results <- readRDS("../data/ts_results.rds")
```

## STL Decomposition

The STL (Seasonal and Trend decomposition using Loess) method separates the 
PM2.5 time series into three components:

- **Trend**: Long-run direction (declining due to policy)
- **Seasonal**: Annual cycle (winter peaks)
- **Remainder**: Short-term variability (weather events, episodes)

```{r stl-plot, fig.cap="STL decomposition of daily PM2.5 at KC1 (Kings College London)."}
stl_df <- ts_results$stl_fit

p1 <- ggplot(stl_df, aes(x = date, y = observed)) +
  geom_line(colour="#5B8DBE", alpha=0.5, linewidth=0.4) +
  labs(title="Observed", y="PM2.5") + theme_minimal(base_size=10)

p2 <- ggplot(stl_df, aes(x = date, y = trend)) +
  geom_line(colour="#FF7043", linewidth=0.8) +
  labs(title="Trend", y=NULL) + theme_minimal(base_size=10)

p3 <- ggplot(stl_df, aes(x = date, y = seasonal)) +
  geom_line(colour="#4CAF50", alpha=0.7, linewidth=0.5) +
  labs(title="Seasonal", y=NULL) + theme_minimal(base_size=10)

p4 <- ggplot(stl_df, aes(x = date, y = remainder)) +
  geom_line(colour="gray40", alpha=0.6, linewidth=0.4) +
  geom_hline(yintercept=0, colour="red", linetype="dashed") +
  labs(title="Remainder", y=NULL) + theme_minimal(base_size=10)

(p1 / p2 / p3 / p4) + plot_annotation(
  title = "STL Decomposition: Daily PM2.5 at KC1"
)
```

## Forecasting Performance

```{r forecast-accuracy}
ts_results$acc_table %>%
  kable(caption = "Forecast accuracy on 90-day test set",
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

---

# Bayesian Hierarchical Model

```{r load-bayes}
bayes_results <- readRDS("../data/bayes_results.rds")
```

## Motivation: Why Bayesian?

The frequentist Negative Binomial GLM provides **point estimates** and 
**confidence intervals** based on asymptotic approximations. The Bayesian 
approach instead provides the **full posterior distribution** of each parameter.

This means we can make direct probabilistic statements:

> *"Given the data and our prior knowledge, there is a 94% probability that PM2.5 
> exposure increases same-day respiratory admissions."*

## Prior Specification

| Parameter | Prior | Rationale |
|---|---|---|
| PM2.5 effect (log RR) | Normal(0, 0.05) | Small positive effect expected |
| Intercept (log scale) | Normal(3, 0.5) | ~20 admissions/day at baseline |
| NB shape (θ) | Half-Cauchy(0, 2.5) | Regularises overdispersion |

## MCMC Convergence

```{r convergence}
bayes_results$rhat_df %>%
  filter(str_detect(parameter, "pm|temp|Intercept|shape")) %>%
  mutate(converged = rhat < 1.01) %>%
  kable(caption = "R-hat convergence diagnostics (< 1.01 = converged)",
        digits = 4) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE) %>%
  column_spec(3, color = ifelse(bayes_results$rhat_df %>% 
    filter(str_detect(parameter, "pm|temp|Intercept|shape")) %>% 
    pull(rhat) < 1.01, "green", "red"))
```

## Posterior Distributions

```{r posterior-plot, fig.cap="Posterior distributions of Rate Ratios. Values to the right of 1 indicate increased admissions risk."}
bayes_results$cred_intervals %>%
  filter(str_detect(parameter, "PM2.5|Temp")) %>%
  ggplot(aes(y = reorder(parameter, median_rr))) +
  geom_vline(xintercept = 1, colour = "gray50", linetype = "dashed") +
  geom_errorbarh(aes(xmin = lower_95, xmax = upper_95), height = 0,
                 linewidth = 2, colour = "#5B8DBE", alpha = 0.4) +
  geom_errorbarh(aes(xmin = lower_80, xmax = upper_80), height = 0,
                 linewidth = 3.5, colour = "#5B8DBE", alpha = 0.7) +
  geom_point(aes(x = median_rr), size = 4, colour = "#FF7043") +
  geom_text(aes(x = upper_95 + 0.002,
                label = paste0("P(harm)=", round(prob_gt_1*100,1),"%")),
            hjust = 0, size = 3.5, colour = "gray40") +
  labs(
    title    = "Bayesian Rate Ratios — Posterior Medians and Credible Intervals",
    subtitle = "Thick bar = 80% CI; thin bar = 95% CI; dot = posterior median",
    x = "Rate Ratio", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  xlim(NA, max(bayes_results$cred_intervals$upper_95) + 0.02)
```

---

# Causal Inference: ULEZ Analysis

```{r load-causal}
causal_results <- readRDS("../data/causal_results.rds")
```

## Design: Interrupted Time Series

**The challenge**: Pollution was already declining before the ULEZ expansion (2021). 
A naive before/after comparison would overestimate the ULEZ effect.

**The solution**: Interrupted Time Series (ITS) models the **pre-existing trend** 
separately from the **intervention effect**, estimating:

1. **Level change**: Immediate step up/down at the intervention date
2. **Slope change**: Change in the rate of decline

## ITS Visualisation

```{r its-plot, fig.cap="ITS analysis: observed vs counterfactual PM2.5 at KC1."}
its_data <- causal_results$its_plot_data

ggplot(its_data) +
  geom_point(aes(x = date, y = pm2.5), alpha = 0.12, size = 0.6, colour = "gray40") +
  geom_line(aes(x = date, y = fitted), colour = "#5B8DBE", linewidth = 1) +
  geom_line(aes(x = date, y = counterfactual),
            colour = "#FF7043", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = as.Date("2021-10-25"), colour = "red", linewidth = 1) +
  annotate("text", x = as.Date("2021-11-10"), y = 42,
           label = "ULEZ", colour = "red", size = 3.5) +
  annotate("text", x = as.Date("2022-06-01"), y = 28,
           label = "Counterfactual\n(no ULEZ)", colour = "#FF7043", size = 3.5) +
  labs(title = "ITS: Observed vs Counterfactual PM2.5 (KC1)",
       subtitle = "Blue = fitted model; Orange dashed = predicted without ULEZ intervention",
       x = NULL, y = "PM2.5 (µg/m³)") +
  theme_minimal(base_size = 12)
```

## Causal Estimates

```{r causal-summary}
causal_results$causal_summary %>%
  kable(caption = "Summary of causal effect estimates") %>%
  kable_styling(bootstrap_options = c("striped","hover"))
```

---

# Machine Learning Comparison

```{r load-ml}
ml_results <- readRDS("../data/ml_results.rds")
```

## Model Comparison

```{r ml-comparison-table}
ml_results$comparison_table %>%
  kable(caption = "All models compared on 90-day test set (lower RMSE = better)",
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  row_spec(1, bold = TRUE, background = "#e8f4f8")
```

## Variable Importance (Random Forest)

```{r rf-importance, fig.cap="Top 15 most important features for next-day PM2.5 prediction (Random Forest)."}
ml_results$rf_importance %>%
  ggplot(aes(x = importance, y = reorder(variable, importance))) +
  geom_col(fill = "#5B8DBE", alpha = 0.85) +
  labs(title = "Random Forest: Variable Importance",
       subtitle = "Recent lag features dominate — PM2.5 has strong autocorrelation",
       x = "Importance", y = NULL) +
  theme_minimal(base_size = 12)
```

## Key Insight: Statistical vs ML

ML models excel at:

- Capturing **non-linear** relationships (e.g., threshold effects)
- Using **many features** simultaneously without manual selection
- **Interaction effects** (e.g., pollution × temperature)

Statistical models (ARIMA, GLM) excel at:

- **Interpretability** — each coefficient has a clear meaning
- **Uncertainty quantification** — principled confidence/credible intervals
- **Extrapolation** — can reason outside training data range
- **Small samples** — ML typically needs more data

---

# Conclusions

```{r conclusions-table, echo=FALSE}
data.frame(
  Finding = c(
    "Seasonal pattern",
    "Trend",
    "Health impact",
    "ULEZ causal effect",
    "Best forecaster",
    "Best predictor overall"
  ),
  Result = c(
    "PM2.5 is 30–40% higher in winter than summer",
    "Declining at ~0.5 µg/m³/year across all sites",
    "1.5% increase in admissions per 10 µg/m³ PM2.5 (Bayesian NB model)",
    "Immediate ~3 µg/m³ reduction at inner-city sites (ITS)",
    "ARIMA — simple, interpretable, competitive accuracy",
    "XGBoost — marginally lower RMSE but black-box"
  )
) %>%
  kable(caption = "Summary of Key Findings") %>%
  kable_styling(bootstrap_options = c("striped","hover"))
```

## Limitations

1. **Health data is synthetic** — generated with published dose-response parameters, 
   not actual NHS records (which are not freely available)
2. **Single city** — London results may not generalise to other urban areas
3. **No individual-level data** — ecological association cannot prove causation at 
   the individual level (ecological fallacy)
4. **ULEZ confounders** — COVID recovery coincided with ULEZ; our ITS controls for 
   trend but other concurrent policies may confound the estimate

## Future Work

- Spatial hierarchical models across all UK cities
- Distributed lag non-linear models (DLNM) for more flexible lag structures
- Ensemble forecasting combining ARIMA, XGBoost, and Bayesian methods
- Individual-level linkage of GP records with pollution monitoring

---

# Appendix: Technical Environment

```{r session-info}
sessionInfo()
```
